### 2014-06-30 patchy-selection-paper.tex
added refs, minor edits to appendices

The convergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes \citep{zhen2012parallel}. 
Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
sometimes over deep time scales \citep{deephomologypapers}. 
Convergence on a molecular level also suggests that the path 
accessible to adaptation is sometimes relatively constrained, 
as would be the case if only a few changes confer the advantagous phenotype 
and are sufficiently free of deleterious pleiotropic consequences.

@@

One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{turner2010serpentine,mimulus};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches. 
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes \citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is primarily dark on a number of black lava flows separated by much lighter soil \citep{dice1940ecologic}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible from the dark pigmentation adaptation on different outcrops \citep{nachman2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}
and in lizards \citep{whitesands}.

@@

To investigate the impact of the various approximations surrounding the mutational influx,
we carried out numerical calculations of a fine-grained, discrete model.
These are shown, and described, in figure~\ref{fig:prob_estab_calcs}.

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-selection-plots.R
        \includegraphics{prob-establishment}
    \end{center}
    \caption{The probability of establishment of a new mutant allele as a function of distance from the edge of the region where it is beneficial,
    in an abrupt transition (left) and a gradual transition (right).
    The allele is deleterious to the left and beneficial to the right (with selection coefficients $\mp 0.05$ respecively);
    and the selection coefficient interpolates linearly between these in the central shaded region.
    The number of offspring is Poisson.
    The underlying landscape has demes every 15km and a migration rate of 0.5 between neighboring demes;
    the probability was found by numerically solving the equation for the probability of establishment of a multi-type branching process.
    \label{fig:prob_estab_calcs}
    }
\end{figure}

@@

One route to the ``equlibrium frequency'' of the allele outside the range where it is advantageous is as follows;
see \citet{slatkin1973geneflow} or \citet{barton1987establishment,pollak1966survival} for other arguments in this case, 
or \citet{etheridge2000introduction} or \citet{dawson1993measurevalued} for general theory used below.
Suppose that the population is composed of a finite number of small demes of equal size $N$ arranged in a regular grid,
and that selection (for or against) the allele is given by the function $s(x)$, with $x$ denoting the spatial location.
Each individual at location $x$ reproduces after a random lifetime,
dying and giving birth in the same deme to a random number of offspring with distribution given by $X$;
each offspring migrates independently to a new location chosen randomly from the distribution given by $x+R$,
where they replaces a randomly chosen individual.
If $x+R$ is outside of the range, then they perish.
Each individual's reproduction time is exponentially distributed, 
the reproduction rate is 1 if it carries the original allele, or with rate $1+s(x)$ if it carries the mutant allele.
Suppose that the number of offspring $X$ has mean $\mu$; the variance of $X$ will not enter into the formula.
Also suppose that the migration displacement $R$ has mean zero and variance $\sigma^2$;
if we are in more than one dimension, we mean that the components of the dispersal distance are uncorrelated
and each have variance $\sigma^2$.
We also assume that the distribution of $X$ makes the associated random walk irreducible and aperiodic.

Let $\Phi^N_t(x)$ be the proportion of mutant alleles present at location $x$ at time $t$,
and $\Phi_t(x)$ the process obtained by taking $N \to \infty$ (which we assume exists).
Denote by $\delta_x$ a single unit at location $x$, so that e.g.~$\Phi^N_t + \delta_x/N$
is the configuration after a mutant allele has been added to location $x$.
For $0\le \phi \le 1$, we also denote by $\bar X_\phi$ the random number of mutant alleles added if $X$ new offspring carrying mutant alleles
replace randomly chosen individuals in a deme where the mutant allele is at frequncy $\phi$ (i.e.~hypergeometric with parameters $(X,\phi)$);
similarly, $\widetilde X_\phi$ is the number lost if the new offspring do not carry the allele (i.e.~hypergeometric with parameters $(X,1-\phi)$).
(We like to think of $\Phi^N_t$ as a measure, but it does not hurt to think of $\Phi^N$ as a vector;
we aren't providing the rigorous justification here.)
Then we know that for any sufficiently nice function $f(\Phi)$ that
\begin{align} \label{eqn:discrete_generator}
  \begin{split} \frac{\partial}{\partial t} \E\left[ f(\Phi^N_t) \right] 
  &= N \sum_x \left\{ \E\left[ (1+s(x+R)) \Phi^N_t(x+R) \left( f\left(\Phi^N_t + \frac{\bar X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right. \\
     & \qquad  \qquad \left. {} + \E\left[ \left(1-\Phi^N_t(x+R)\right) \left( f\left(\Phi^N_t - \frac{\widetilde X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right\}  \end{split} \\
     &= \mu \sum_x \E\left[ \left(\partial_{\phi(x)} f(\Phi_t) \right) \left\{ \Phi_t(x+R) - \Phi_t(x) + s(x+R) \Phi_t(x+R) (1-\Phi_t(x)) \right\} \right] + O\left(\frac{1}{N}\right).
\end{align}
In the final expectation, $R$ and $\Phi$ are independent.
This follows by taking first-order terms in $1/N$ in the Taylor series for $f$, 
and the fact that $\E[\bar X_\phi] = \phi \mu$ and $\E[\widetilde X_\phi] = (1-\phi)\mu$.
We can see two things from this:
First, since this is a first-order differential operator, the limiting stochastic process $\Phi$
is in fact deterministic (check by applying to $f(\Phi) = \Phi(x)^2$ to find the variance).
Second, if we want to rescale space as well to get the usual differential equation, 
we need to choose $\var[R]=\sigma^2$ and $s(x)$ to be of the same, small, order; 
this is another way of seeing that $\sigma/\sqrt{s}$ is the relevant length scale.
More concretely, suppose that the grid size is $\epsilon \to 0$, 
that $\var[R] = (\sigma \epsilon)^2$, and that $s(x)/\epsilon \to \gamma(x)$,
and suppose that $\Phi_t(x)$ is deterministic (and sufficiently differentiable) and let $\xi(t,x) = \Phi_{t/\epsilon}(x)$;
then the previous equation with $f(\Phi) = \Phi(x)$ converges to the familiar form:
\begin{align}
  \label{eqn:derived_fkpp}
  \partial_t \xi(t,x) = \mu \left( \frac{\sigma^2}{2} \sum_{k=1}^d \partial_{x_k}^2 \xi(t,x) + \gamma(x) \xi(t,x) (1-\xi(t,x)) \right) .
\end{align}
Here we have taken first the population size $N \to \infty$ and then the grid size $\epsilon \to 0$;
we could alternatively take both limits together, but not if $\epsilon$ goes to zero too much faster than $N$ grows.
One reason for this is that at finite $N$,
the process $\Phi_t$ is an irreducible, aperiodic finite-state Markov chain with absorbing states at 0 and 1;
therefore, the inevitable outcome is extinction of one type or another,
which is not the regime we want to study.

@@

We are interested in the following case:
\[
    s(x) = \begin{cases}
        +s_b \qquad \mbox{for } |x|<\epsilon \\
        -s_d \qquad \mbox{for } x<-\epsilon \; \mbox{or}\; \epsilon \le x \le R \\
        -s_* \qquad \mbox{for } R < x < R+2\epsilon \\
        -s_d \qquad \mbox{for } x>R+2\epsilon .
    \end{cases}
\]
We can solve the case with $s_* = s_d$, since $\phi'(0) = 0$ (as done implicitly in \citet{slatkin1973geneflow}).
We should be able to argue that if $R$ is large then the value of $s_*$ won't make much difference.

@@

Presumably, this gets the same answer as Monty, who finds that in his solvable cases,
for $x$ in the last interval (which ends in $\infty$), that
\begin{align*}
    \phi(x) &= \frac{3}{2}\left( 1 - \tanh^2 ( x \sqrt{2s} / 2 \sigma + C ) \right) \\
        &\simeq C' \exp\left( - \frac{ x \sqrt{2s} }{ \sigma } \right) .
\end{align*}
% Indeed, as $\phi \to 0$, we have approximately $\phi' = \frac{ \sqrt{s} }{ \sigma } \phi$.  % huh????

@@

- leimu2008metaanalysis: local adaptation not ubiquitous but common; mostly in large population sizes.
- robertson1960theory: dependence of Ne on the degree of adaptation
- lots of local adaptation stuff on plants and conservation genetics; ties into plasticity and range limits.
- but also: Neurospora (ellison2011) and salmonids (fraser2011)
- swamping in Mayr 1963, Animal Species and Evolution?
- hanson1966effects: original observation of minimum patch size, in simulation?
- macnair1991: larger effects more common in e.g. heavy metal sites
- schat1996: silene parallel adaptation to mine tailings
- kaweki: a review.
- fourneir-level2011: map of arabidopsis local adaptation
- reviews for local adapt in marine (sanford2011), trees (savolainen2007)

>>>>

The convergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes \citep{zhen2012parallel}. 
Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
sometimes over deep time scales \citep{deephomologypapers}. 
Convergence on a molecular level also suggests that the path 
accessible to adaptation is sometimes relatively constrained \citep{stern2013genetic}, 
as would be the case if only a few changes confer the advantagous phenotype 
and are sufficiently free of deleterious pleiotropic consequences.

@@

One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{macnair1991evolution,schat1996identical,turner2010serpentine,mimulus};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches. 
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes \citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is primarily dark on a number of black lava flows separated by much lighter soil \citep{dice1940ecologic}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible from the dark pigmentation adaptation on different outcrops \citep{nachman2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}
and in lizards \citep{whitesands}.

@@

To investigate the impact of the various approximations surrounding the mutational influx,
we calculated the probability of establishment for an spatial branching process moving across discrete demes.
The solution was obtained by iterating the multivariate generating function until convergence
(since the function which gives the probability of extinction as a function of location of the progenitor
is an attractive fixed point of the generating function; see \citet{jagers1975branching}).
These solutions are shown, and parameters described, in figure~\ref{fig:prob_estab_calcs}.

@@

\begin{figure}[ht!]
    \begin{center}
        % made in patchy-selection-plots.R
        \includegraphics{prob-establishment}
    \end{center}
    \caption{The probability of establishment of a new mutant allele as a function of distance from the edge of the region where it is beneficial,
    in an abrupt transition (left) and a gradual transition (right).
    The allele is deleterious to the left and beneficial to the right (with selection coefficients $\mp 0.05$ respecively);
    and the selection coefficient interpolates linearly between these in the central hatched region.
    The number of offspring is Poisson.
    The underlying landscape has demes every 15km and a migration rate of 0.5 between neighboring demes;
    the probability was found by numerically solving the equation for the probability of establishment of a multi-type branching process.
    \label{fig:prob_estab_calcs}
    }
\end{figure}

@@

For completeness, and clarity as to the scalings on the relevant parameters,
here we provide a derivation of the differential equations referred to above equation~\eqref{eqn:eqfreq},
and establish the asymptotics given in that equation.
One route to the ``equlibrium frequency'' of the allele outside the range where it is advantageous is as follows;
see \citet{slatkin1973geneflow} or \citet{barton1987establishment,pollak1966survival} for other arguments in this case, 
or \citet{etheridge2000introduction} and/or \citet{dawson1993measurevalued} for general theory used below.
Following is a description of a discrete model (which is nearly the one we simulate from).
Suppose that the population is composed of a finite number of small demes of equal size $N$ arranged in a regular grid,
and that selection (for or against) the allele is given by the function $s(x)$, with $x$ denoting the spatial location.
Each individual at location $x$ reproduces after a random lifetime,
dying, and then leaving behind in the same deme a random number of offspring with distribution given by $X$;
each offspring then migrates independently to a new location chosen randomly from the distribution given by $x+R$,
where she replaces a randomly chosen individual.
If $x+R$ is outside of the range, then she perishes.
Each individual's time until reproduction is exponentially distributed:
the reproduction rate is 1 if it carries the original allele, or with rate $1+s(x)$ if it carries the mutant allele.
Suppose that the number of offspring $X$ has mean $\mu$; the variance of $X$ will not enter into the formula.
Also suppose that the migration displacement $R$ has mean zero and variance $\sigma^2$;
if we are in more than one dimension, we mean that the components of the dispersal distance are uncorrelated
and each have variance $\sigma^2$.
We also assume that the distribution of $R$ makes the associated random walk irreducible and aperiodic.

Let $\Phi^N_t(x)$ be the proportion of mutant alleles present at location $x$ at time $t$,
and $\Phi_t(x)$ the process obtained by taking $N \to \infty$ (which we assume exists).
Denote by $\delta_x$ a single unit at location $x$, so that e.g.~$\Phi^N_t + \delta_x/N$
is the configuration after a mutant allele has been added to location $x$.
For $0\le \phi \le 1$, we also denote by $\bar X_\phi$ the random number of mutant alleles added if $X$ new offspring carrying mutant alleles
replace randomly chosen individuals in a deme where the mutant allele is at frequncy $\phi$ (i.e.~hypergeometric with parameters $(X,\phi)$);
similarly, $\widetilde X_\phi$ is the number lost if the new offspring do not carry the allele (i.e.~hypergeometric with parameters $(X,1-\phi)$).
(We like to think of $\Phi^N_t$ as a measure, but it does not hurt to think of $\Phi^N$ as a vector;
we aren't providing the rigorous justification here.)
Then we know that for any sufficiently nice function $f(\Phi)$ that
\begin{align} \label{eqn:discrete_generator}
  \begin{split} \frac{\partial}{\partial t} \E\left[ f(\Phi^N_t) \right] 
  &= N \sum_x \left\{ \E\left[ (1+s(x+R)) \Phi^N_t(x+R) \left( f\left(\Phi^N_t + \frac{\bar X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right. \\
     & \qquad  \qquad \left. {} + \E\left[ \left(1-\Phi^N_t(x+R)\right) \left( f\left(\Phi^N_t - \frac{\widetilde X_{\Phi_t(x+R)}}{N}\delta_{x}\right) - f(\Phi^N_t) \right) \right] \right\}  \end{split} \\
     &= \mu \sum_x \E\left[ \left(\partial_{\phi(x)} f(\Phi_t) \right) \left\{ \Phi_t(x+R) - \Phi_t(x) + s(x+R) \Phi_t(x+R) (1-\Phi_t(x)) \right\} \right] + O\left(\frac{1}{N}\right).
\end{align}
In the final expectation, $R$ and $\Phi$ are independent.
This follows by taking first-order terms in $1/N$ in the Taylor series for $f$, 
and the fact that $\E[\bar X_\phi] = \phi \mu$ and $\E[\widetilde X_\phi] = (1-\phi)\mu$.
We can see two things from this:
First, since this is a first-order differential operator, the limiting stochastic process $\Phi$
is in fact deterministic (check by applying to $f(\Phi) = \Phi(x)^2$ to find the variance).
Second, if we want to rescale space as well to get the usual differential equation, 
we need to choose $\var[R]=\sigma^2$ and $s(x)$ to be of the same, small, order; 
this is another way of seeing that $\sigma/\sqrt{s}$ is the relevant length scale \citep[as noted by][]{slatkin1973geneflow}.
More concretely, suppose that the grid size is $\epsilon \to 0$, 
that $\var[R] = (\sigma \epsilon)^2$, and that $s(x)/\epsilon \to \gamma(x)$,
and suppose that $\Phi_t(x)$ is deterministic (and sufficiently differentiable) and let $\xi(t,x) = \Phi_{t/\epsilon}(x)$;
then the previous equation with $f(\Phi) = \Phi(x)$ converges to the familiar form:
\begin{align}
  \label{eqn:derived_fkpp}
  \partial_t \xi(t,x) = \mu \left( \frac{\sigma^2}{2} \sum_{k=1}^d \partial_{x_k}^2 \xi(t,x) + \gamma(x) \xi(t,x) (1-\xi(t,x)) \right) .
\end{align}
Here we have taken first the population size $N \to \infty$ and then the grid size $\epsilon \to 0$;
we could alternatively take both limits together, but not if $\epsilon$ goes to zero too much faster than $N$ grows.
One reason for this is that at finite $N$,
the process $\Phi_t$ is an irreducible, aperiodic finite-state Markov chain with absorbing states at 0 and 1;
therefore, the inevitable outcome is extinction of one type or another,
which is not the regime we want to study.

@@

We are interested in the following case:
\[
    s(x) = \begin{cases}
        +s_b \qquad \mbox{for } |x|<\epsilon \\
        -s_d \qquad \mbox{for } x<-\epsilon \; \mbox{or}\; \epsilon \le x \le R \\
        -s_* \qquad \mbox{for } R < x < R+2\epsilon \\
        -s_d \qquad \mbox{for } x>R+2\epsilon .
    \end{cases}
\]
We can solve the case with $s_* = s_d$, since $\phi'(0) = 0$ (as done implicitly in \citet{slatkin1973geneflow}).
It should be possible to argue that if $R$ is large then the value of $s_*$ won't make much difference.

@@

% Presumably, this gets the same answer as \citep{slatkin1973geneflow}, who finds that in the solvable cases,
% for $x$ in the last interval (which ends in $\infty$), that
% \begin{align*}
%     \phi(x) &= \frac{3}{2}\left( 1 - \tanh^2 ( x \sqrt{2s} / 2 \sigma + C ) \right) \\
%         &\simeq C' \exp\left( - \frac{ x \sqrt{2s} }{ \sigma } \right) .
% \end{align*}

@@

- leimu2008metaanalysis: local adaptation not ubiquitous but common; mostly in large population sizes.
- robertson1960theory: dependence of the degree of adaptation on Ne 
- lots of local adaptation stuff on plants and conservation genetics; ties into plasticity and range limits.
- but also: Neurospora (ellison2011) and salmonids (fraser2011)
- swamping in Mayr 1963, Animal Species and Evolution?
- hanson1966effects: original observation of minimum patch size, in simulation?
- fourneir-level2011: map of arabidopsis local adaptation
- reviews for local adapt in marine (sanford2011), trees (savolainen2007)
